---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Introduction

This report addresses a set of requirement included herein: the enumered section of this report corresponds to the "review criteria". All of the R code needed to reproduce the results (numbers, plots, etc.) are included in this report.

**Load libraries**
```{r}
#load all required libraries
library(lattice)
library(magrittr)
library(dplyr)
```

## Loading and preprocessing the data
**1. Code for reading in the dataset and/or processing the data**
```{r, echo=TRUE}
#read data (see activity.zip) from current directory
activity <- read.csv("activity.csv", header = TRUE, sep = ",", na.strings = "NA")
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
**2. Histogram of the total number of steps taken each day**
```{r, echo=TRUE}
#calculate total number of steps taken per day
TotalStepsPerDay1 <- activity %>% select(date, steps) %>% group_by(date) %>% summarise_each(funs(sum))

#generate the histogram
hist(TotalStepsPerDay1$steps, breaks = 10, xlab = "Interval", main = "Total number of steps")
```

**3. Mean and median number of steps taken each day**
```{r, echo=TRUE}
meanPerDay1 <- mean(TotalStepsPerDay1$steps, na.rm = TRUE)
medianPerDay1 <- median(TotalStepsPerDay1$steps, na.rm = TRUE)
```
* Mean: `r meanPerDay1`
* Median: `r medianPerDay1`

## What is the average daily activity pattern?
**4. Time series plot of the average number of steps taken**
```{r, echo=TRUE}
#calculate average number of steps by interval
AvgStepsByInterval <- activity %>% select(steps, interval) %>% group_by(interval) %>% summarise_each(funs(mean(., na.rm = TRUE)))

#generate the timeseries plot
with(AvgStepsByInterval, plot(x = interval, y = steps, type = "l", xlab = "Interval", main = "Average number of steps"))
```

**5. The 5-minute interval that, on average, contains the maximum number of steps**
```{r, echo=TRUE}
AvgStepsByInterval[AvgStepsByInterval$steps==max(AvgStepsByInterval$steps),]
```

## Imputing missing values
**6. Code to describe and show a strategy for inputting missing data**
```{r, echo=TRUE}
sum(is.na(activity$steps)==TRUE) #total number of missing values
```

Replace missing value with the average of a given interval:
```{r, echo=TRUE}
activity$steps[is.na(activity$steps)] <- AvgStepsByInterval$steps 
```

**7. Histogram of the total number of steps taken each day after missing values are inputted**
```{r, echo=TRUE}
#calculate total number of steps per day
TotalStepsPerDay2 <- activity %>% select(date, steps) %>% group_by(date) %>% summarise_each(funs(sum))

#generate the histogram
hist(TotalStepsPerDay2$steps, breaks = 10, xlab = "Interval", main = "Total number of steps") #plot histogram
```

In addition, subsequent to inputting the missing value we can compare the mean and median values (see review criteria 3); which are expectingly equal.

```{r, echo=TRUE}
meanPerDay2 <- mean(TotalStepsPerDay2$steps, na.rm = TRUE)
medianPerDay2 <- median(TotalStepsPerDay2$steps, na.rm = TRUE)
```
* Mean: `r meanPerDay2`
* Median: `r medianPerDay2`


## Are there differences in activity patterns between weekdays and weekends?
**8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends**
```{r, echo=TRUE}
#transform data accordingly
weekend <- c("Saturday", "Sunday")
activity$w <- factor((weekdays(activity$date) %in% weekend), levels = c("TRUE", "FALSE"), labels=c("weekend", "weekday"))
activity1 <- activity %>% select(steps,interval,w) %>% group_by(interval, w) %>% summarise_each(funs(mean))

#generate the plot
xyplot(steps ~ interval | w, data = activity1, type = "l", layout=c(1,2), main = "Average number of steps")
```







